Portfolios of Subgraph Isomorphism Algorithms

### 2015-12-15 paper/paper.tex
proofread up to section 6

\subsection{Considered Existing Algorithms}

We have selected a set of algorithms with complementary performance. These algorithms will compose our portfolio in the per-instance selection approach described in Section 5. 
\begin{itemize}
\item \VFtwo{}~\cite{Cordella:2004} performs a weak filtering that it is especially fast on
    trivially satisfiable instances;
\item \LAD{}~\cite{Solnon:2010} combines two strong but expensive filtering techniques
    (All-Different Generalized Arc Consistency and Locally All-Different);
\item \Glasgow{}~\cite{McCreesh:2015} does expensive preprocessing based on path length
    invariant properties to generate additional constraints, followed by weaker filtering
    (forward-checking, and a heuristic All-Different propagator which can miss deletions) and
    conflict-directed backjumping during search.
\end{itemize}

\noindent The \Glasgow algorithm has a parameter, which controls the lengths of paths used when
reasoning about non-adjacent vertices.  In experiments reported by McCreesh and
Prosser~\cite{McCreesh:2015}, the choice of paths of length 3 was used as a reasonable compromise---longer
paths lead to prohibitively expensive preprocessing on larger, denser instances. This is often not
the best choice on an instance by instance basis: sometimes path-based reasoning gives no benefit at
all, sometimes considering only paths of length 2 suffices, occasionally paths of length 4 are
helpful, and even looking at paths of length 3 is relatively expensive on some graphs. We thus consider all
four choices, naming these variants \GlasgowOne through \GlasgowFour.

\subsection{New algorithms}

We also introduce two new variants of \LAD. The first, called \IncompleteLAD, is a weaker
filtering which is applied once, without performing a backtracking search, and very quickly
detects inconsistencies on many instances: for each pattern node $u$, we check that there exists at
least one target node $v$ such that for each neighbor $u'$ of $u$ there exists a different neighbor
$v'$ of $v$ such that the degree of $u'$ is smaller than or equal to the degree of $v'$.
\IncompleteLAD is an incomplete algorithm as it can only detect inconsistency of some instances: when
it does not detect inconsistency, the instance may either be satisfiable or not. As a counterpart of
this incompleteness, it is very fast: its time complexity is ${\cal O}(n_p(n_t+e_t))$.

@@

We consider a large benchmark of 5725 instances, which are available in a simple text
format\footnote{\url{http://liris.cnrs.fr/csolnon/SIP.html}}. These instances are grouped into 12
classes.

\begin{itemize}
\item Class 1 contains randomly generated scale-free graphs~\cite{constraints10}.
\item Classes 2 and 3 contain various kinds of graphs selected by Larrosa and Valiente~\cite{LV02}:
    class 2 contains small instances generated from the 50 first graphs, and class 3 contains
    larger instances where pattern graphs are coming from class 2, and target graphs are coming from
    the next 50 graphs.
\item Classes 4 to 8 contain randomly generated graphs coming from a database of graphs commonly
    used for benchmarking subgraph isomorphism algorithms~\cite{GraphDatabase1,GraphDatabase2}: bounded-degree graphs for classes 3 and 4, regular meshes
    for classes 5 and 6, and random graphs with uniform edge probabilities for class 7. All of these
    instances are satisfiable.
\item Classes 9 and 10 contain instances coming from segmented images~\cite{pr15,cviu11}.
\item Class 11 contains instances coming from meshes modeling 3D objects~\cite{cviu11}.
\item Class 12 contains random graph instances chosen to be close to the satisfiable-unsatisfiable
    phase transition---these instances are expected to be particularly challenging, despite their
    small size.
\end{itemize}

Note that Classes 3 and 12 were not considered in the previous experimental study by McCreesh and
Prosser~\cite{McCreesh:2015}.  This set of instances is also much larger than that of Battiti and
Mascia~\cite{battiti-mascia07}, who were the first to propose algorithm portfolios for subgraph
isomorphism problems.  Battiti and Mascia only considered a pure parallel portfolio consisting of
two solvers, and without a selection mechanism. Their problem set consisted entirely of easy
satisfiable instances, which is where this style of approach has most success.

@@

We measured runtimes on machines with Intel Xeon E5-2640 v2 CPUs and 64GBytes RAM, running
Scientific Linux 6.5. We used the C++ implementation of the \Glasgow algorithm~\cite{McCreesh:2015},
the C implementation of \LAD{}~\cite{Solnon:2010}, and the VFLib C
implementation of \VFtwo{}~\cite{Cordella:2004}. Software was compiled using GCC 4.9. Each problem instance was run with a
timeout of $10^8$ milliseconds (a little over a day).

\subsection{Results} \label{expComp}

\begin{figure}[t]
\includegraphics[width=\textwidth]{courbe.pdf}
\caption{Evolution of the number of solved instances with respect to CPU time, for the 8 algorithms
of the portfolio, and a Virtual Best Solver (VBS).\label{expTimeGraph}}
\end{figure}

Figure \ref{expTimeGraph} displays the evolution of the number of instances solved with respect to
CPU time. It shows us that the best solver depends on the time limit considered. \IncompleteLAD is
able to solve easy unsatisfiable instances very quickly, in a few milliseconds. Hence, for time
limits lower than \SI{5}{\ms}, the best solver is \IncompleteLAD. However, it is not able to solve harder
unsatisfiable instances, nor can it solve satisfiable instances.

\PathLAD and \GlasgowOne outperform \IncompleteLAD when considering longer time limits: \PathLAD is the
best solver for time limits larger than \SI{5}{\ms} and lower than \SI{40}{\ms}, and \GlasgowOne is the best solver
for time limits larger than \SI{40}{\ms} and lower than \SI{3000}{\ms}.

For CPU time limits larger than \SI{3000}{\ms}, the best solver becomes \GlasgowTwo.  As we
increase the CPU time limit, variants of \Glasgow with longer paths (\GlasgowThree and \GlasgowFour)
become better. This is what we expect, as more reasoning is expensive, but potentially increases the
reduction of the search space. In particular, the relative performance of Glasgow variants is
monotone in the time it takes to solve the instance---\GlasgowOne is best at first, but gets worse as
the time to solve the instance increases until \GlasgowTwo is better. After that point, \GlasgowOne is
never better than \GlasgowTwo again. The same is true for path lengths 3 and 4.

@@

Furthermore, the Virtual Best Solver (VBS), which considers the best algorithm for each instance
separately, obtains much better results, showing us that the algorithms have complementary
performance. The difference between VBS and single best is more particularly important for rather
small CPU time limits (smaller than \SI{1000}{\ms}). In many applications, it is important to have the
fastest possible algorithm. For example, in pattern
recognition~\cite{pr15,cviu11} and chemical~\cite{Giugno:2013} applications, we often have to solve the subgraph isomorphism problem repeatedly
for a very large number of graphs (in order to find a pattern image or molecule in a large database
of target images or compounds, for example), so having an algorithm that is able to solve an
instance in \SI{100}{\ms} instead of \SI{1000}{\ms} makes a big difference.  Therefore, it is important to select
the best algorithm for each instance, even if the instance is an easy one.

Table~\ref{expClass} shows us that we cannot simply select algorithms based upon which class from
which instances are coming (unlike in SAT, where this kind of approach can be reasonable). For all
classes, there are always at least two algorithms which are the best for at least one instance of
the class. In particular, for classes 2 and 3, each algorithm is the best for at least one instance
(except \GlasgowFour for class 3).

@@

Experimental results reported in Section~\ref{expComp} have shown us that \IncompleteLAD is very fast
(\SI{7}{\ms} on average) and is able to solve 1919 instances very quickly. Therefore, we first run
\IncompleteLAD: if the instance is detected to be unsatisfiable, then we do not process it further.
Otherwise, we record as features the number of candidate couples removed by \IncompleteLAD in absolute
terms, as a percentage, and the minimum and maximum on a per-variable basis.

\VFtwo is also able to solve many easy instances very quickly: among the 3806 instances which are not
solved by \IncompleteLAD, 1470 are solved by \VFtwo in less than 50 milliseconds. Therefore, after
running \IncompleteLAD, we run the \VFtwo solver for \SI{50}{\ms}. This solves easy instances without
the overhead of running algorithm selection and avoids potentially making incorrect solver choices.
After running \IncompleteLAD and \VFtwo for \SI{50}{\ms}, we are left with 2336 hard instances that we
consider for algorithm selection.

@@

If presolving does not give us a solution, we extract additional features. For both the pattern and
the target graph, we consider some basic graph properties, which may be computed very quickly:

\begin{itemize}
    \item The number of vertices and edges.
    \item The density---we expect that some kinds of filtering might be expensive and ineffective on
        dense graphs.
    \item How many loops (self-adjacent vertices) the graph contains---since loops must be mapped to
        loops, this could have a strong effect on how easy an instance is.
    \item The mean and maximum degrees, and whether or not every vertex has the same degree. (The
        degree-based invariants used by \LAD and \Glasgow do nothing at the top of search if every
        vertex has the same degree.)
    \item Whether or not the graph is connected.
    \item The mean and maximum distances between all pairs of vertices (if nearly all vertices are
        close together, path-based reasoning is likely to be ineffective). We also consider the
        proportion of vertex pairs which are distance at least 2, 3 and 4 apart.
\end{itemize}

@@

To decide which solver to run on the remaining instances, we use \LLAMA~\cite{kotthoff_llama_2013}.
\LLAMA supports the most common algorithm selection approaches used in the literature. We performed a
set of preliminary experiments to determine the approach that works best in this case.

@@

\LLAMA's pairwise regression approach with random forest regression gave the best performance. The
idea is very similar to the pairwise classification used by Xu et al.~\cite{xu_satzilla_2008}. For
each pair of algorithms in our portfolio, we train a model that predicts the performance difference
between them. That is, if the first algorithm is better than the second, the difference is positive,
otherwise negative. The algorithm with the highest cumulative performance difference, i.e.\ the most
positive difference over all other algorithms, is chosen to be run.

@@

Table~\ref{expTimeTable} displays the number of instances solved at different CPU time limits,
ranging from 1 to $10^8$ milliseconds. It shows us that the best single solver depends on the time
limit considered. Simple approaches like \IncompleteLAD and \VFtwo are able to solve easy instances
very quickly, in a few milliseconds. However, they are not able to solve harder instances:
\IncompleteLAD is an incomplete approach which can only detect rather trivial inconsistencies; \VFtwo
performs a basic backtracking search and is very fast on easy instances, because it does not compute
expensive invariants. Classical \LAD, \PathLAD, and the \Glasgow algorithms solve more instances than \VFtwo
and \IncompleteLAD when considering longer time limits: \SI{10}{\ms} for \LAD and \PathLAD and
\GlasgowOne, and \SI{100}{\ms} (resp. \SI{1000}{\ms} and \SI{10000}{\ms}) for \GlasgowTwo (resp.\ 3 and
4). The Virtual Best Solver (VBS), which considers the best algorithm for each instance separately,
obtains much better results, showing us that the algorithms have complementary performance. The
difference between VBS and single best (column $\delta$ of Table \ref{expTimeTable}) is more
particularly important for rather small CPU time limits. In particular, VBS solves 693 more
instances than the single best (\GlasgowOne) when the limit is \SI{100}{\ms}. In many applications,
it is important to have the fastest possible algorithm. For example, in pattern recognition
applications, we often have to solve subgraph isomorphism problems for a very large number of graphs
(in order to find a pattern image in a large database of target images, for example), so that having
an algorithm that is able to solve an instance in \SI{100}{\ms} instead of \SI{1000}{\ms} makes a
big difference. Therefore, it is important to select the best algorithm for each instance.

>>>>

We consider six algorithms from the literature and propose two novel ones.

\subsection{Algorithms from the Literature}

We have selected a set of algorithms with complementary performance. These
algorithms will compose our portfolio in the per-instance selection approach
described in Section~\ref{sec:algsel}.
\begin{itemize}
\item \VFtwo{}~\cite{Cordella:2004} performs weak filtering that it is especially fast on
    trivially satisfiable instances;
\item \LAD{}~\cite{Solnon:2010} combines two strong but expensive filtering techniques
    (All-Different Generalized Arc Consistency and Locally All-Different);
\item \Glasgow{}~\cite{McCreesh:2015} does expensive preprocessing based on path length
    invariant properties to generate additional constraints, followed by weaker filtering
    (forward-checking, and a heuristic All-Different propagator which can miss deletions) and
    conflict-directed backjumping during search.
\end{itemize}

\noindent The \Glasgow algorithm has a parameter, which controls the lengths of paths used when
reasoning about non-adjacent vertices.  In experiments reported by McCreesh and
Prosser~\cite{McCreesh:2015}, the choice of paths of length 3 was used as a reasonable compromise---longer
paths lead to prohibitively expensive preprocessing on larger, denser instances. This is often not
the best choice on an instance by instance basis: sometimes path-based reasoning gives no benefit at
all, sometimes considering only paths of length 2 suffices, occasionally paths of length 4 are
helpful, and even looking at paths of length 3 is relatively expensive on some graphs. We thus consider all
lengths up to 4, naming these variants \GlasgowOne through \GlasgowFour.

\subsection{Novel Algorithms}

We introduce two new variants of \LAD. The first, called \IncompleteLAD, does weaker
filtering which is applied once, without performing a backtracking search, and very quickly
detects inconsistencies on many instances: for each pattern node $u$, we check that there exists at
least one target node $v$ such that for each neighbor $u'$ of $u$ there exists a different neighbor
$v'$ of $v$ such that the degree of $u'$ is smaller than or equal to the degree of $v'$.
\IncompleteLAD is an incomplete algorithm that checks a sufficient, but not
necessary, condition for inconsistency: when it does not detect inconsistency,
the instance may still be unsatisfiable. Its main benefit is that it runs very
fast: its time complexity is ${\cal O}(n_p(n_t+e_t))$.

@@

We consider a large benchmark set of 5725 instances, which are available in a simple text
format\footnote{\url{http://liris.cnrs.fr/csolnon/SIP.html}}. These instances are grouped into 12
classes.

\begin{itemize}
\item Class 1 contains randomly generated scale-free graphs~\cite{constraints10}.
\item Classes 2 and 3 contain various kinds of graphs selected by Larrosa and Valiente~\cite{LV02}:
    class 2 contains small instances generated from the 50 first graphs of class
    1 and class 3 contains
    larger instances with pattern graphs from class 2 and target graphs from
    the next 50 graphs.
\item Classes 4 to 8 contain randomly generated graphs from a database of graphs commonly
    used for benchmarking subgraph isomorphism algorithms~\cite{GraphDatabase1,GraphDatabase2}: bounded-degree graphs for classes 4 and 5, regular meshes
    for classes 6 and 7, and random graphs with uniform edge probabilities for class 8. All of these
    instances are satisfiable.
\item Classes 9 and 10 contain instances from segmented images~\cite{pr15,cviu11}.
\item Class 11 contains instances from meshes modeling 3D objects~\cite{cviu11}.
\item Class 12 contains random graph instances chosen to be close to the satisfiable-unsatisfiable
    phase transition---these instances are expected to be particularly challenging, despite their
    small size.
\end{itemize}

Note that Classes 3 and 12 were not considered in the previous experimental study by McCreesh and
Prosser~\cite{McCreesh:2015}.  This set of instances is also much larger than that of Battiti and
Mascia~\cite{battiti-mascia07}, who were the first to propose algorithm portfolios for subgraph
isomorphism problems.  Battiti and Mascia only considered a pure parallel portfolio consisting of
two solvers, and without a selection mechanism. Their problem set consisted entirely of easy
satisfiable instances.

@@

We measured runtimes on machines with Intel Xeon E5-2640 v2 CPUs and 64GBytes RAM, running
Scientific Linux 6.5. We used the C++ implementation of the \Glasgow algorithm~\cite{McCreesh:2015},
the C implementation of \LAD{}~\cite{Solnon:2010}, and the VFLib C
implementation of \VFtwo{}~\cite{Cordella:2004}. Software was compiled using GCC 4.9. Each problem instance was run with a
timeout of $10^8$ milliseconds ($\approx$ 27.8 hours).

\subsection{Results}\label{expComp}

\begin{figure}[t]
\includegraphics[width=\textwidth]{courbe.pdf}
\caption{Number of solved instances over CPU time for the eight algorithms we
consider in this paper, and the virtual best solver (VBS) that shows the best
solver on an instance by instance basis.\label{expTimeGraph}}
\end{figure}

Figure~\ref{expTimeGraph} displays the evolution of the number of instances solved with respect to
CPU time. It shows us that the best solver depends on the time limit considered. \IncompleteLAD is
able to solve easy unsatisfiable instances very quickly, in a few milliseconds. Hence, for time
limits lower than \SI{5}{\ms}, the best solver is \IncompleteLAD. However, it is not able to solve harder
unsatisfiable instances, nor can it solve satisfiable instances.

\PathLAD and \GlasgowOne outperform \IncompleteLAD for longer time limits: \PathLAD is the
best solver for time limits larger than \SI{5}{\ms} and lower than \SI{40}{\ms}, and \GlasgowOne is the best solver
for time limits larger than \SI{40}{\ms} and lower than \SI{3000}{\ms}.

\GlasgowTwo becomes the best solver for time limits larger than \SI{3000}{\ms}.
As we
increase the CPU time limit, variants of \Glasgow with longer paths (\GlasgowThree and \GlasgowFour)
become better. This is what we expect, as more reasoning is expensive, but
increases the potential
reduction of the search space. In particular, the relative performance of \Glasgow variants is
monotone in the time it takes to solve the instance---\GlasgowOne is best at first, but gets worse as
the time to solve the instance increases until \GlasgowTwo is better. After that point, \GlasgowOne is
never better than \GlasgowTwo again. The same is true for path lengths 3 and 4.

@@

Furthermore, the virtual best solver (VBS), which considers the best algorithm
for each instance separately, obtains much better results, showing us that the
algorithms have complementary performance. The difference between VBS and single
best is particularly pronounced for CPU time limits less than \SI{1000}{\ms}. In
many applications, it is important to have the fastest possible algorithm even
if the absolute differences in CPU time are small. For example, in pattern
recognition~\cite{pr15,cviu11} and chemical~\cite{Giugno:2013} applications, we
often have to solve the subgraph isomorphism problem repeatedly for a very large
number of graphs (in order to find a pattern image or molecule in a large
database of target images or compounds, for example), so having an algorithm
that is able to solve an instance in \SI{100}{\ms} instead of \SI{1000}{\ms}
makes a big difference.  Therefore, it is important to select the best algorithm
for each instance, even if the instance is an easy one.

Table~\ref{expClass} shows us that we cannot simply select algorithms based on
the instance class. For all
classes, there are always at least two algorithms which are the best for at least one instance of
the class. In particular, for classes 2 and 3, each algorithm is the best for at least one instance
(except \GlasgowFour for class 3).

@@

Experimental results reported in Section~\ref{expComp} have shown us that \IncompleteLAD is very fast
(\SI{7}{\ms} on average) and is able to solve 1919 instances from our benchmark
set very quickly. Therefore, we first run
\IncompleteLAD: if unsatisfiability is detected, we do not need to process it further.
Otherwise, we record the number of candidate pairs removed and as a percentage
of all pairs and the minimum and maximum values on a per-variable basis. The
recorded values are then used as features for algorithm selection.

\VFtwo is also able to solve many easy instances very quickly: among the 3806 instances which are not
solved by \IncompleteLAD, 1470 are solved by \VFtwo in less than 50 milliseconds. Therefore, after
running \IncompleteLAD, we run the \VFtwo solver for \SI{50}{\ms}. This solves easy instances without
the overhead of running algorithm selection and avoids potentially making incorrect solver choices.

After the presolving step, we are left with 2336 hard instances that we consider
for algorithm selection.

@@

If presolving does not give us a solution, we extract additional features. For both the pattern and
the target graph, we consider some basic graph properties, which can be computed
very quickly,
\begin{itemize}
    \item the number of vertices and edges;
    \item the density---we expect that some kinds of filtering might be expensive and ineffective on
        dense graphs;
    \item how many loops (self-adjacent vertices) the graph contains---since loops must be mapped to
        loops, this could have a strong effect on how easy an instance is;
    \item the mean and maximum degrees, and whether or not every vertex has the same degree (the
        degree-based invariants used by \LAD and \Glasgow do nothing at the top of search if every
        vertex has the same degree);
    \item whether or not the graph is connected;
    \item the mean and maximum distances between all pairs of vertices (if nearly all vertices are
        close together, path-based reasoning is likely to be ineffective) and
        the proportion of vertex pairs which are at least 2, 3 and 4 apart.
\end{itemize}

@@

We use \LLAMA~\cite{kotthoff_llama_2013} to build our algorithm selection model.
\LLAMA supports the most common algorithm selection approaches used in the literature. We performed a
set of preliminary experiments to determine the approach that works best here.

@@

\LLAMA's pairwise regression approach with random forest regression gave the best performance. The
idea is very similar to the pairwise classification models used by Xu et al.~\cite{xu_satzilla_2008}. For
each pair of algorithms in our portfolio, we train a model that predicts the performance difference
between them. If the first algorithm is better than the second, the difference is positive,
otherwise negative. The algorithm with the highest cumulative performance difference, i.e.\ the most
positive difference over all other algorithms, is chosen to be run.

@@

Table~\ref{expTimeTable} displays the number of instances solved at different CPU time limits,
ranging from 1 to $10^8$ milliseconds. It shows us that the best single solver depends on the time
limit considered. Simple approaches like \IncompleteLAD and \VFtwo are able to solve easy instances
very quickly, in a few milliseconds. However, they are not able to solve harder instances:
\IncompleteLAD is an incomplete approach which can only detect rather trivial inconsistencies; \VFtwo
performs a basic backtracking search and is very fast on easy instances, because it does not compute
expensive invariants. Classical \LAD, \PathLAD, and the \Glasgow algorithms solve more instances than \VFtwo
and \IncompleteLAD when considering longer time limits: \SI{10}{\ms} for \LAD and \PathLAD and
\GlasgowOne, and \SI{100}{\ms} (resp. \SI{1000}{\ms} and \SI{10000}{\ms}) for \GlasgowTwo (resp.\ 3 and
4). The virtual best solver obtains much better results, showing us that the algorithms have complementary performance. The
difference between VBS and single best (column $\delta$ of Table \ref{expTimeTable}) is more
particularly important for rather small CPU time limits. In particular, VBS solves 693 more
instances than the single best (\GlasgowOne) when the limit is \SI{100}{\ms}. In many applications,
it is important to have the fastest possible algorithm. For example, in pattern recognition
applications, we often have to solve subgraph isomorphism problems for a very large number of graphs
(in order to find a pattern image in a large database of target images, for example), so that having
an algorithm that is able to solve an instance in \SI{100}{\ms} instead of \SI{1000}{\ms} makes a
big difference. Therefore, it is important to select the best algorithm for each instance.
